Allocation Removal by Partial Evaluation in a Tracing JIT

### 2010-10-06 talk/pepm2011/paper.tex
first go at integrating the operational semantics. not completely happy.

\begin{figure}
\begin{center}
\includegraphics[scale=0.7]{figures/obj-lifetime.pdf}
\end{center}

@@

XXX want some sort of pseudo-code

>>>>

\begin{figure}
\begin{center}
\includegraphics[scale=0.6]{figures/obj-lifetime.pdf}
\end{center}

@@

\begin{figure*}
\begin{center}
\begin{tabular}{lcccc}
\emph{new} & ${\displaystyle \frac{l\,\mathrm{fresh}}{v=\mathrm{new}(T),E,H\overset{\mathrm{run}}{\Longrightarrow}E\left[v\mapsto l\right],H\left[l\mapsto\left(T,\mathrm{null},\mathrm{null}\right)\right]}}$ & ~~~ &
\emph{guard} & ${\displaystyle \frac{\mathrm{type}(H(E(v))=T}{\mathrm{guard}(v,T),E,H\overset{\mathrm{run}}{\Longrightarrow}E,H}}$\tabularnewline[3em]
\emph{get} & ${\displaystyle \frac{\,}{v=\mathrm{get}(u,F),E,H\overset{\mathrm{run}}{\Longrightarrow}E\left[v\mapsto H\left(E\left(u\right)\right)_{F}\right],H}}$ & ~~~ &
& ${\displaystyle \frac{\mathrm{type}(H(E(v))\neq T}{\mathrm{guard}(v,T),E,H\overset{\mathrm{run}}{\Longrightarrow}\bot,\bot}}$\tabularnewline[3em]
\emph{set} & ${\displaystyle \frac{\,}{\mathrm{set}\left(v,F,u\right),E,H\overset{\mathrm{run}}{\Longrightarrow}E,H\left[E\left(v\right)\mapsto\left(H\left(E\left(v\right)\right)!_{F}E(u)\right)\right]}}$ & ~~~ &
& \tabularnewline[4em]
\end{tabular}

\begin{minipage}[b]{7 cm}
\emph{Object Domains:}
$$\setlength\arraycolsep{0.1em}
 \begin{array}{rcll}
    u,v,w & \in & V & \mathrm{\ variables\ in\ trace}\\
    T & \in & \mathfrak{T} & \mathrm{\ runtime\ types}\\
    F & \in & \left\{ L,R\right\} & \mathrm{\ fields\ of\ objects}\\
    l & \in & L & \mathrm{\ locations\ on\ heap}
 \end{array}
$$
\end{minipage}
\begin{minipage}[b]{5 cm}
\emph{Semantic Values:}
$$\setlength\arraycolsep{0.1em}
 \begin{array}{rcll}
    E & \in & V\rightharpoonup L & \mathrm{\ Environment}\\
    H & \in & L\rightharpoonup\mathfrak{T}\times(L\cup \{ \mathrm{null} \})\times (L\cup \{ \mathrm{null}\} ) & \mathrm{\ Heap}\\
    \\
    \\
 \end{array}
$$
\end{minipage}
\end{center}
\label{fig:semantics}
\caption{The Operational Semantics of Simplified Traces}
\end{figure*}

In this section we want to give a formal description of the semantics of the
traces and of the optimizer and liken the optimization to partial evaluation.
We concentrate on the operations for manipulating dynamically allocated objects,
as those are the only ones that are actually optimized. Without loss of
generality we also consider only objects with two fields in this section.

Traces are lists of operations. The operations considered here are new (to make
a new object), get (to read a field out of an object), set (to write a field
into an object) and guard (to check the type of an object). The values of all
variables are locations (i.e.~pointers). Locations are mapped to objects, which
are represented by triples of a type $T$, and two locations that represent the
fields of the object. When a new object is created, the fields are initialized
to null, but we require that they are immediately initialized to a real
location, otherwise the trace is malformed.

We use some abbreviations when dealing with object triples. To read the type of
an object, $\mathrm{type}((T,l_1,l_2))=T$ is used. Reading a field $F$ from an
object is written $(T,l_1,l_2)_F$ which either returns $l_1$ if $F=L$ or $l_2$
if $F=R$. To set field $F$ to a new location $l$, we use the notation
$(T,l_1,l_2)!_Fl$, which yields a new triple $(T,l,l_2)$ if $F=L$ or a new
triple $(T,l_1,l)$ if $F=R$.

Figure~\ref{fig:semantics} shows the operational semantics for traces. The
interpreter formalized there executes one operation at a time. Its state is
represented by an environment and a heap, which are potentially changed by the
execution of an operation. The environment is a partial function from variables
to locations and the heap is a partial function from locations to objects. Note
that a variable can never be null in the environment, otherwise the trace would
be malformed. We use some syntax for changing partial functions: $E[v\mapsto l]$
denotes the environment which is just like $E$, but maps $v$ to $l$.

The new operation creates a new object $(T,\mathrm{null},\mathrm{null})$, on the
heap under a fresh location $l$ and adds the result variable to the environment,
mapping to the new location $l$.

The get operation reads a field $F$ out of an object and adds the result
variable to the environment, mapping to the read location. The heap is
unchanged.

The set operation changes field $F$ of an object stored at the location that
variable $v$ maps to. The new value of the field is the location in variable
$u$. The environment is unchanged.

The guard operation is used to check whether the object stored at the location
that variable $v$ maps to is of type $T$. If that is the case, then execution
continues without changing heap and environment. Otherwise, execution is
stopped.

\subsection{Optimizing Traces}
\label{sub:formalopt}

% subsection Optimizing Traces (end)

\begin{figure*}
\begin{center}
\fbox{
\begin{tabular}{lc}
\emph{new} & ${\displaystyle \frac{v^{*}\,\mathrm{fresh}}{v=\mathrm{new}(T),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \,\right\rangle ,E\left[v\mapsto v^{*}\right],S\left[v^{*}\mapsto\left(T,\mathrm{null,null}\right)\right]}}$\tabularnewline[3em]
\emph{get} & ${\displaystyle \frac{E(u)\in\mathrm{dom}(S)}{v=\mathrm{get}(u,F),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \,\right\rangle ,E\left[v\mapsto S(E(u))_{F}\right],S}}$\tabularnewline[3em]
 & ${\displaystyle \frac{E(u)\notin\mathrm{dom}(S)\, v^{*}\,\mathrm{fresh}}{v=\mathrm{get}(u,F),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle v^{*}=\mathrm{get}(E(u),F)\right\rangle ,E\left[v\mapsto v^{*}\right],S}}$\tabularnewline[3em]
\emph{set} & ${\displaystyle \frac{E(v)\in\mathrm{dom}(S)}{\mathrm{set}\left(v,F,u\right),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \,\right\rangle ,E,S\left[E\left(v\right)\mapsto\left(S(E(v))!_{F}E(u)\right)\right]}}$\tabularnewline[3em]
 & ${\displaystyle \frac{E(v)\notin\mathrm{dom}\left(S\right),\,\left(E(v),S\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops},S^{\prime}\right)}{\mathrm{set}\left(v,F,u\right),E,S\overset{\mathrm{opt}}{\Longrightarrow}\mathrm{ops}::\left\langle \mathrm{set}\left(E(v),F,E(u)\right)\right\rangle ,E,S^{\prime}}}$\tabularnewline[3em]
\emph{guard} & ${\displaystyle \frac{E(v)\in\mathrm{dom}(S),\,\mathrm{type}(S(E(v)))=T}{\mathrm{guard}(v,T),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \,\right\rangle ,E,S}}$\tabularnewline[3em]
 & ${\displaystyle \frac{E(v)\in\mathrm{dom}(S),\,\mathrm{type}(S(E(v)))\neq T,\,\left(E(v),S\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops},S^{\prime}\right)}{\mathrm{guard}(v,T),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \mathrm{guard}(E\left(v\right),T)\right\rangle ,E,S^{\prime}}}$\tabularnewline[3em]
 & ${\displaystyle \frac{E(v)\notin\mathrm{dom}(S)}{\mathrm{guard}(v,T),E,S\overset{\mathrm{opt}}{\Longrightarrow}\left\langle \mathrm{guard}(E\left(v\right),T)\right\rangle ,E,S}}$\tabularnewline[3em]
\emph{lifting} & ${\displaystyle \frac{v^{*}\notin\mathrm{dom}(S)}{v^{*},S\overset{\mathrm{lift}}{\Longrightarrow}\left\langle \,\right\rangle ,S}}$\tabularnewline[3em]
 & ${\displaystyle \frac{v^{*}\in\mathrm{dom}(S),\,\left(v^{*},S\setminus\left\{ v^{*}\mapsto S\left(v^{*}\right)\right\} \right)\overset{\mathrm{liftfields}}{\Longrightarrow}\left(\mathrm{ops},S^{\prime}\right)}{v^{*},S\overset{\mathrm{lift}}{\Longrightarrow}\left\langle v^{*}=\mathrm{new}\left(T\right)\right\rangle ::ops,S^{\prime}}}$\tabularnewline[3em]
 & ${\displaystyle \frac{\left(S\left(v^{*}\right)_{L},S\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops}_{L},S^{\prime}\right),\,\left(S\left(v^{*}\right)_{R},S^{\prime}\right)\overset{\mathrm{lift}}{\Longrightarrow}\left(\mathrm{ops}_{R},S^{\prime\prime}\right)}{v^{*},S\overset{\mathrm{liftfields}}{=\!=\!\Longrightarrow}\mathrm{ops}_{L}::ops_{R}::\left\langle \mathrm{set}\left(v^{*},L,S\left(v^{*}\right)_{L}\right),\,\mathrm{set}\left(v^{*},R,S\left(v^{*}\right)_{R}\right)\right\rangle ,S^{\prime}}}$\tabularnewline
\end{tabular}
}
\end{center}
\label{fig:optimization}
\caption{Optimization Rules}
\end{figure*}
